---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---
## Load libraries
```{r Libraries, results="hide"}
library(dplyr)

```

## Loading and preprocessing the data
```{r Loading Data, echo = TRUE}
act <- read.csv("activity.csv", stringsAsFactors = FALSE)
summary(act)
```

Since there is a date value, we will convert that into a date class.
```{r Convert Date, echo = TRUE}
act_date <- as.Date(act$date)

```



There are many NA's in the Steps column, we'll convert those to zero to avoid issues with plotting.
```{r Remove NA, echo = TRUE}
act$steps <- as.integer(act$steps)
```

## What is mean total number of steps taken per day?

First we group the steps by day.

```{r daily step histogram, echo = TRUE}
act_day <- summarize(group_by(act,date), 
                     steps = sum(steps, na.rm = TRUE))

print(act_day)

hist(act_day$steps)
```

Now we calculate the Mean and Median
```{r average daily steps, echo = TRUE}

daily_steps_mean <- round(mean(act_day$steps))
print(paste("Average Daily Steps is", daily_steps_mean))

daily_steps_med <- round(median(act_day$steps))
print(paste("Median Daily Steps is", daily_steps_med))
```


## What is the average daily activity pattern?
First we make a time series chart of all 5 min5ute intervals and calculate the average steps taken every 5 min5utes.
```{r Interval, echo=TRUE}
min5 <- summarise(group_by(act,interval), steps = mean(steps, na.rm = TRUE))

plot(min5$interval, min5$steps, type = "l", main = "Average Steps for each interval", xlab = "Interval", ylab = "Steps")


max_pos <- match(max(min5$steps),min5$steps)
print( paste0("The highest steps interval is ", min5$interval[max_pos] ) )
```

## Inputing missing values
First we check for how many incomplete rows there are: 
```{r Missing Values, echo = TRUE}
act2 <- act
for(i in names(act2)){
    print(paste0("Column ", i, 
                 " has ",
                 sum(is.na(act2[i])) ,
                 " NA's"))
}
```
There are 2304 incomplete rows all in the steps column.  We will attempt to substitute and fill in those empty cells with the mean across all intervals.

```{r Subbing Values, echo = TRUE}
    pos <- 0
    for( i in act2$steps){
        pos <- pos + 1
        if(is.na(i)){
            inter <- act2$interval[pos]
            inter_pos <- which(act2$interval == inter)
            int_mean <- mean(act2$steps[inter_pos], na.rm = TRUE)
            act2$steps[pos] <- round(int_mean)
        }
    }
```

Now we look at the new stats for the filled in dataframe:
```{r Checking filled in data, echo = TRUE}
act2_day <- summarize(group_by(act2,date), 
                     steps = sum(steps, na.rm = TRUE))

daily_steps_mean2 <- round(mean(act2_day$steps))
print(paste("Average Daily Steps is with filled in values", daily_steps_mean2))

daily_steps_med2 <- round(median(act2_day$steps))
print(paste("Median Daily Steps with filled in values is", daily_steps_med2))

hist(act2_day$steps)

```

After filling the values, it is clear from the historgrams that the daily step values have shifted toward the median.

## Are there differences in activity patterns between weekdays and weekends?

We first label what day of the week each date is and break the data down inbetween weekend and weekday

```{r Weekday converstion, echo = TRUE}
wk <- weekdays(as.Date(act2$date), abbreviate = TRUE)
act3 <- cbind(act2,wk)
wkend <- subset(act3, act3$wk == "Sun"|act3$wk == "Sat")
wkday <- subset(act3, act3$wk != "Sun" & act3$wk != "Sat")
```

Now we plot the 2 data sets

```{r Plot weekday vs weekend, echo = TRUE}
wkday_sum <- summarise(group_by(wkday,interval), steps = mean(steps, na.rm = TRUE))

plot(wkday_sum$interval, wkday_sum$steps, type = "l", main = "Weekday vs Weekend Step Pattern", xlab = "Interval", ylab = "Steps", ylim = c(0,300), col = "blue")

wkend_sum <- summarise(group_by(wkend,interval), steps = mean(steps, na.rm = TRUE))

lines(wkend_sum$interval, wkend_sum$steps, col = "green")
legend("topright", c("Weekday","Weekend"),col = c("blue","green"), bg = "gray", cex = .75, lty = 1, bty = "n")
```

The overlayed graphs show there is more steps taken during weekday mornings, and then some more in the evening while the weekends have a more even spread of walking throughout the day.